\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-11-08 writing/manuscript.Rmd
added some refs

\newpage <!-- start main body on a new page -->

@@

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its probability to survive and reproduce. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019], or even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will change their behaviours in response to changes in local resource abundance. While some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020]. However, movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @lucherini_habitat_1996; @relyea_home_2000], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with larger decreases when resources are scarce [e.g., @herfindal_prey_2005; @nilsen_can_2005]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago various scientists, inlcuding @lindstedt_home_1986 and @reiss_scaling_1988. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.

@@

### Expected resource abundance, $\mathbb E(R)$

The expectation of a random variable (i.e. its mean or average, indicated here as ) is the value one can *expect* the random variable to take. $\mathbb E(R) = \mu$ to indicate the expectation of the random variable $R$. The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\mathbb E(R)$ is still useful for estimating the average long-term resource abundance. When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\mathbb E(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\mathbb E(R) = \mu(t)$.

@@

### Variance in resource abundance, $\mathbb V(R)$

A random variable's variance is a measure of its unpredictability or variability. We use  $\mathbb V(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$. For instance, while one may consider $R$ to vary within years, we define $\mathbb V(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation. We refer to known differences or expected spatiotemporal changes as **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable once an animal visits the valley. In contrast, whether next year will be a good or bad year for berries is stochastic.

@@

To understand how movement can be viewed as a stochastic process, imagine an animal that starts at its den and moves to collect food and other resources, and it returns to its den once it has collected enough resources for the day. If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another. However, we can use the animal's positional data to estimate the amount of time it spends exploring before returning "home" (or the time required for the position to "decay" back to the average position), which is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$. Similarly, we can define $\tau_v$ to be amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity. $\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter. Thus, by modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for a variety properties of tracking data which are problematic for models that assume errors to be independent while also leveraging such properties to estimate important movement parameters. Accounting for autocorrelation in movement data (which is present in most modern datasets) avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as path length and average (or instantaneous) speed [@noonan_scale-insensitive_2019].

@@

The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Thus, here we quantify animal's spatial needs using their positional variance (indicated as $\mathbb V(\text{position})$), instead. ***explain the concept of positional variance*** [@refs].

Although range-resident animals will generally have smaller $\mathbb V(\text{position})$ than migratory and nomadic animals, $\mathbb V(\text{position})$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\mathbb V(\text{position})$ determines the scale and size of an animal's spatial needs but not its behavior. Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different kinds of movement. Animals with a $\tau_p$ shorter than a year are likely range-resident, while animals with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year). In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it is unlikely to return to locations it visited in the past.

@@

Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement.

@@

@@

```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\mathbb V(\\text{position})$) of an animal as a function of mean resource abundance ($\\mathbb E(R)$). We expect low values of $\\mathbb E(R)$ to result in high $\\mathbb V(\\text{position})$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic. As $\\mathbb E(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive. Note that the relationship between $\\mathbb E(R)$ and $\\mathbb V(\\text{position})$ cannot be linear because it would require $\\mathbb V(\\text{position})$ to be negative for high values of $\\mathbb E(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @ref]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [e.g., @nilsen_can_2005; @ex; see figure \@ref(fig:mu-hyp)]. Low values of $\mathbb E(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp). As $\mathbb E(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

@@

```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\mathbb V(\\text{position})$) of an animal as a function of resource unpredictability ($\\mathbb V(R)$). We expect low values of $\\mathbb V(R)$ to result in low $\\mathbb V(\\text{position})$ as animals are able to depend on somewhat predictable resources. As $\\mathbb V(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration. Note that the relationship between $\\mathbb V(R)$ and $\\mathbb V(\\text{position})$ cannot be linear because it would require $\\mathbb V(\\text{position})$ to be negative for low values of $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
```

\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\mathbb V(R)$ has received far less attention [but see @lucherini_habitat_1996; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\mathbb V(R)$ can change strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018]. Environmental variability, including extreme events [@logares_black_2012], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs. A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.

Simlarly to the effect of $\mathbb E(R)$, we expect $\mathbb V(R)$ to have a nonlinear effect on $\mathbb V(\text{position})$. When $\mathbb V(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\mathbb V(R)$ to have a noticeable effect on $\mathbb V(\text{position})$. However, as resources become additionally unpredictable, we expect spatial needs to increase superlinearly. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors, including nomadism, dispersal, and migration (figure \@ref(fig:s2-hyp)).

The effects of $\mathbb V(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs, so the degree to which $R$ varies will likely affects each species differently. Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with a larger $\mathbb V(\text{position})$ may not detect such differences. @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity. Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic. In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions. Similarly, @nilsen_can_2005 found that the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased with seasonality, while the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality. The work by @nilsen_can_2005 also demonstrates that the effect of $\mathbb E(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\mathbb V(R)$, so it is important to account for interaction effects between the two.

* predation is more stochastic than herbibory; predators need to match more cycles, depend on fewer food sources that are able to move $\rightarrow$ higher effect of $\mathbb V(R)$

@@

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with constant variance (after accounting for changes in mean abundance). (a.) Although resource abundance ($R$, raster fill) varies over time and space, the overall amount of variance (dot color) remains constant. (b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$. (c.) Simulated animal movement at a time of lowest $\\mathbb E(R)$ and average $\\mathbb V(R)$. (d.) Simulated animal movement at a time of peak $\\mathbb E(R)$ and average $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

\noindent Consider the simplistic scenario where $\mathbb E(R) = \mu$ changes over time but $\mathbb V(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that the resource abundance follows a Gamma distribution with a mean that changes over time (i.e., the expected value is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d). However, if $\mathbb E(R) = \mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019]. However, since $\mathbb E(R)$ does not vary predictably over space, it will not be possible to predict the location of locations with high or low $R$.

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case [@refs; but see @refs]. Most often, $\mathbb E(R)$ will vary over time, and animals' behaviors will change in response. Additionally, when data availability is sufficiently high to produce precise estimates of $\mathbb V(R)$, the effect of $\mathbb V(R)$ on animals' $\mathbb V(\text{position})$ should be estimated before assuming the effect is negligible.

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with changing variance (even after accounting for changes in mean abundance). (a.) Although resource abundance ($R$, brown-green fill) varies over time and space, variance (dot color) is lowest at the beginning and end of the observational period and highest when $R$ peaks. (b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$. (c.) Simulated animal movement at a time of lowest $\\mathbb E(R)$ and $\\mathbb V(R)$. (d.) Simulated animal movement at a time of peak $\\mathbb E(R)$ and $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```

>>>>

<!-- add custom commands for formulae -->

\newcommand{\e}{\mathbb E}
\newcommand{\var}{\mathbb V}
\newcommand{\vpos}{\mathbb V(\text{position})}

<!-- have table of contents on pages of its own -->

@@

<!-- start main body on a new page -->
\newpage

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its probability to survive and reproduce. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019], or even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will change their behaviors in response to changes in local resource abundance. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @lucherini_habitat_1996; @relyea_home_2000], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., @herfindal_prey_2005; @nilsen_can_2005]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists, including @lindstedt_home_1986 and @reiss_scaling_1988. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.

@@

### Expected resource abundance, $\e(R)$

The expectation of a random variable (i.e. its mean or average, indicated here as ) is the value one can *expect* the random variable to take. $\e(R) = \mu$ to indicate the expectation of the random variable $R$. The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\e(R)$ is still useful for estimating the average long-term resource abundance. When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\e(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\e(R) = \mu(t)$.

@@

### Variance in resource abundance, $\var(R)$

A random variable's variance is a measure of its unpredictability or variability. We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$. For instance, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation. We refer to known differences or expected spatiotemporal changes as **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable once an animal visits the valley. In contrast, whether next year will be a good or bad year for berries is stochastic.

@@

To understand how movement can be viewed as a stochastic process, imagine an animal that starts at its den and moves to collect food and other resources, and it returns to its den once it has collected enough resources for the day. If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another. However, we can use the animal's positional data to estimate the amount of time it spends exploring before returning "home" (or the time required for the position to "decay" back to the average position), which is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$. Similarly, we can define $\tau_v$ to be amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity. $\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter. Thus, by modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for a variety properties of tracking data which are problematic for models that assume errors to be independent while also leveraging such properties to estimate important movement parameters. Accounting for autocorrelation in movement data (which is present in most modern datasets) avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as path length and average (or instantaneous) speed [@noonan_scale-insensitive_2019].

@@

The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the variance in position of animals that are not range-resident [@refs].

Although range-resident animals will generally have smaller $\vpos$ than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's spatial needs but not its behavior. Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different kinds of movement. Animals with a $\tau_p$ shorter than a year are likely range-resident, while animals with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year). In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it is unlikely to return to locations it visited in the past.

@@

Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement. Animals may temporarily change their behavior in response to many causes, including predators [@ref], competitors [@ref], weather [@ref], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].

@@

@boratynski_energetic_2020: animals' hr is limited by BMI and maximum metabolic rate, which can be accounted for by accounting for body mass [@noonan_effects_2020]

@arechavala-lopez_common_2019: octopuses (*Octopus vulgaris*) congregate in human-altered coastal habitats

@mitsuhashi_home_2018: urban raccoon dogs (*Nyctereutes procyonoides*) have a smaller HR than rural ones

@cronk_home_2021: yellow mongoose (*Cynictis penicillata*) in urban regions have smaller HRs

@@

```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\var(\\text{position})$) of an animal as a function of mean resource abundance ($\\e(R)$). We expect low values of $\\e(R)$ to result in high $\\var(\\text{position})$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic. As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive. Note that the relationship between $\\e(R)$ and $\\var(\\text{position})$ cannot be linear because it would require $\\var(\\text{position})$ to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @ref]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [e.g., @nilsen_can_2005; @ex; see figure \@ref(fig:mu-hyp)]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

@@

```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\var(\\text{position})$) of an animal as a function of resource unpredictability ($\\var(R)$). We expect low values of $\\var(R)$ to result in low $\\var(\\text{position})$ as animals are able to depend on somewhat predictable resources. As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration. Note that the relationship between $\\var(R)$ and $\\var(\\text{position})$ cannot be linear because it would require $\\var(\\text{position})$ to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
```

Although it generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].

\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\var(R)$ has received far less attention [but see @lucherini_habitat_1996; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\var(R)$ can change strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018]. Environmental variability, including extreme events [@logares_black_2012], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs. A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.

Simlarly to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on $\vpos$. When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect on $\vpos$. However, as resources become additionally unpredictable, we expect spatial needs to increase superlinearly. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors, including nomadism, dispersal, and migration (figure \@ref(fig:s2-hyp)).

The effects of $\var(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs, so the degree to which $R$ varies will likely affects each species differently. Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with a larger $\vpos$ may not detect such differences. @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity. Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic. In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions. Similarly, @nilsen_can_2005 found that the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased with seasonality, while the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality. The work by @nilsen_can_2005 also demonstrates that the effect of $\e(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\var(R)$, so it is important to account for interaction effects between the two.

* predation is more stochastic than herbibory; predators need to match more cycles, depend on fewer food sources that are able to move $\rightarrow$ higher effect of $\var(R)$

@@

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with constant variance (after accounting for changes in mean abundance). (a.) Although resource abundance ($R$, raster fill) varies over time and space, the overall amount of variance (dot color) remains constant. (b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$. (c.) Simulated animal movement at a time of lowest $\\e(R)$ and average $\\var(R)$. (d.) Simulated animal movement at a time of peak $\\e(R)$ and average $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

\noindent Consider the simplistic scenario where $\e(R) = \mu$ changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that the resource abundance follows a Gamma distribution with a mean that changes over time (i.e., the expected value is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d). However, if $\e(R) = \mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019]. However, since $\e(R)$ does not vary predictably over space, it will not be possible to predict the location of locations with high or low $R$.

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case [@refs; but see @refs]. Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response. Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' $\vpos$ should be estimated before assuming the effect is negligible.

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with changing variance (even after accounting for changes in mean abundance). (a.) Although resource abundance ($R$, brown-green fill) varies over time and space, variance (dot color) is lowest at the beginning and end of the observational period and highest when $R$ peaks. (b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$. (c.) Simulated animal movement at a time of lowest $\\e(R)$ and $\\var(R)$. (d.) Simulated animal movement at a time of peak $\\e(R)$ and $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```

When animals are not guaranteed that the resources they find during one visit will be there next time (figure \@ref(fig:multipanel-changing-var)a), $\var(R)$ will have an appreciable effect on the location's favourableness. Since $\var(R)$ is not constant over time, we can now define it as a function of time using the notation $\var(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b). In this scenario, both the expected $R$ and its predictability change over time, so accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).
